Lack of Evidence?: Academic Probation Does Not Affect Student Dropout and Subsequent Performance

### 2021-04-20 model_test_script.Rmd
Commit for Results

>>>>

```{r}
gpa_camp2 <- lm(GPA_year2 ~ status + dist_from_cut + gender + first_language + highHS*study_campus, data = clean_cut)
summary(gpa_camp2)
```

```{r}
dropout_camp2 <- glm(left_school ~ status + dist_from_cut + gender + highHS*study_campus + first_language ,family = "binomial", data = clean_cut)
summary(dropout_camp2)
```

```{r}
grad_camp2 <- glm(gradin4 ~ status + dist_from_cut + gender + highHS*study_campus + first_language ,family = "binomial", data = clean_cut)
summary(grad_camp2)
```

### 2021-04-20 outputs/paper/paper.Rmd
Commit for Results

---
title: "Academic Probation Increases Student Dropout While Improving Remaining Student Performance"
subtitle: "Revisiting the Ability, Gender, and Performance Standards: Evidence from Academic Probation (Lindo, Sanders, and Oreopoulos 2010)"
author: "Hong Shi"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: " Many universities use academic probation as a wake-up call to ensure currently enrolled students to achieve minimum academic standards. This paper examines the causal impact on first-year students in a Canadian univesity after being placed on academic probation in a sharp regression discontinuity design. The experiment results indicate that academic probation increases the student dropout, improves GPAs of the remaining students and negatively affects graduation rates. There are also heterogeneous responses across students' high school performance, gender, native language, campus and program of study towards academic probation. #High Level Importance \\par\\textbf{Keywords:} Regression Discontinuity Design, Academic Probation, Dropoff, Student Performance, Graduation, Causality, Simpson's Paradox"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
toc: FALSE

@@

&nbsp; &nbsp; This paper examines the causal impact on first-year university students after being placed on academic probation by exploiting the discontinuous probation GPA cutoff in a sharp regression discontinuity design (RDD). The data comes form a large Canadian university with three individual campuses. It is a replication of `“Ability, Gender, and Performance Standards: Evidence from Academic Probation."` [@cite_original] with adjustment on the bandwidth (i.e., student within certain grade points of the academic probation cutoff). The results are consistent with Lindo's findings that being placed on probation at the end of the first year increases the student dropout, improves the GPAs of the remaining students and negatively affects graduation rates (Note: May adjust the result after applying the new bandwidth.). Besides, students with different high school academic performance, gender or native language react differently to academic probation. Furthermore, I also find heterogeneous responses to academic probation across campus that is not mentioned in the original paper (Note: May adjust the result after running the model.).

&nbsp; &nbsp; The rest of this paper is organized as follows: In Data Section, I would first explain the domain background about the academic probation. Next, I introduce the data used in the original paper. And then, I would describe the Quasi-experiment design rationale and also raise concerns about the bandwidth used in the original paper. In Model section, I elaborate on the sharp RDD and presents the model in this experiment. I would address assumptions about RDD as well. In Results section, I show the impact of academic probation on student dropout, subsequent performance and graduation with the adjusted bandwidth. In Discussion section, I would first summarize the experiment findings, and then elaborate on the causality of the experiment, address external and internal validity and also Simpson's Paradox (also possibly Berkson's Paradox) regarding the campus variation. Finally, I conclude with some limitations and suggest future directions of this experiment. ^[Code and data are available at the GitHub repo: https://github.com/honn-ishinn/RDD_academic_probation.]

# Data

&nbsp; &nbsp; The data comes from a large Canadian university with three individual campuses: one central campus (Campus 1) and two smaller satellite campuses (Campus 2 and Campus 3) from 1996 to 2005. I analyzed it using R [@citeR], and packages `tidyverse` [@cite_tidyverse],  `here` [@cite_here], `rdrobust` [@cite_rdrobust], `rdd`[@cite_rdd], `rdpower`[@cite_rdpower],  `stargazer`[@cite_stargazer], `janitor` [@cite_janitor], `httr` [@cite_httr], `xml2` [@cite_xml2]. I used packages `bookdown` [@cite_bookdown], `kableExtra` [@cite_kableextra], `ggrepel` [@cite_ggrepel], `finalfit` [@cite_finalfit], `broom` [@cite_broom] to format the document and referenced `Impact Evaluation in Practice` [@cite_impact] to evaluate this experiment.

@@

```{r, include=FALSE}
clean_data <- read.csv(here("inputs/data/cleaned_original.csv"))
```

@@

4. Restrict students having their academic standing evaluated at the end of Year 1 ($98\%$ of the remaining sample).

@@

&nbsp; &nbsp; In this study, since the RDD only provides the most accurate estimates around the probation cutoff where treatment and comparison are most similar [@cite_impact], it is critical to determine the bandwidth(i.e., students within certain grade points of the probation cutoff score) to maintain the balance of observed characteristics of students around the probation cutoff. In Lindo's original paper, they choose the bandwidth of 0.6 and the following table \@ref(tab:lindo) shows the summary statistics of student observable characteristics within 0.6 GPA.

@@

```{r lindo, tab.cap="Summary Statistics of Students Within 0.6 GPA of Probation Cutoff", echo = FALSE, message = FALSE, warning=FALSE}
kable(combine3[,c(4,1,2,3)],caption = "Summary Statistics of Students Within 0.6 GPA of Probation Cutoff",
       booktabs = TRUE, linesep = "") %>% 
  kable_styling(latex_options = "hold_position")
```
&nbsp; &nbsp; According to @cite_GPA_determinants, highschool grade is one of the key determinants of university academic success. Thus, highschool grades could somehow reveal students' study potential in the university, so it is preferable to balance student highschool performance to obtain a more accurate estimate of academic probation. However, the t-test \@ref(tab:ttesthslin) conducted on highschool grade percentile suggests a large difference on highschool performance between student observations in good standing and on academic probation within Lindo's 0.6 GPA bandwidth. In addition, the study on the course load on academic success [@cite_credits] implies that course credit taken by students could reflect student academic ability. Therefore, another t-test \@ref(tab:ttestcrlin) is conducted on total course credits attempted^[Failing a course will not earn course credit but count as an attempted credit] in Year 1 between students in good standing and on probation. The result suggests a difference in the course workload between good standing and probation students within Lindo's 0.6 GPA bandwidth. As a result, the bandwidth that Lindo used in the original paper may include students with different academic potentials, reducing the local estimation accuracy of placing first-year students on academic probation.

@@

&nbsp; &nbsp; The following table \@ref(tab:gradingtheme) shows the grading theme of this Canadian university^[This is the current grading theme of this university as of 2021. The grading theme between 1996 and 2005 might be different and could not be retrieved on the university website. While the current grading theme may still serve as a reference of this study]. By applying Lindo's bandwidth of 0.6 GPA, observations barely above the probation cutoff roughly covers students within 60-66 average grade percentage (1.5-2.1 GPA in Campus 1 and 2, 1.6-2.2 GPA in Campus 3), and observations barely below the probation cutoff roughly covers students within 53-59 average grade percentage (0.9-1.5 GPA in Campus 1 and 2, 1.0-1.6 GPA in Campus 3). By converting the grade point value into grade percentage, it might become more obvious that there exists academic performance difference between students in good standing with 66 average grade percentage and students on probation with 53 average grade percentage. So I would argue that the bandwidth used in Lindo's is too wide to maintain similar characteristics, particularly academic potentials, between the probation and good standing students.

@@

&nbsp; &nbsp; The power of a test is the probability that the test will reject the test null hypothesis to detect a correct alternative hypothesis [@cite_stats_tests]. In this study, as the length of the bandwidth increases, more students are included within the bandwidth. The increase in the number of student raises the power to detect the effect on placing students on probation so that the effect is less likely to happen by chance rather than by the probation cutoff rule. Therefore, I use the `rdpower`[@cite_rdpower] package to calculate the sufficient sample size required (i.e., effect size) to detect the probation effect on (1) Student dropout, (2) Second year GPA and (3) Graduation rate by Year 4. The result as shown below (Table \@ref(tab:powersample)) suggests that the bandwidth of 0.2 GPA with 4106 students (1745 on probation, 2361 in good standing) could raise sufficient statistical power. Further adjustment on the bandwidth might either unbalance student characteristics (Increase bandwidth) or decrease the power of the study (Decrease bandwidth). So I would apply the bandwidth of 0.2 GPA to examine the impact of academic probation in the following session.

@@

# Model

@@

&nbsp; &nbsp; The university policy ensures that all first year students below the probation cutoff will be placed into academic probation (as shown in Figure \@ref(fig:statsassign))^[The original data shows university administrative errors in data reporting that falsely place students either above the cutoff on probation or below the cutoff in good standing (98 out of 44362 student observations). These dirty data are removed from this study.]. So the discontinuity around the cutoff is "sharp".

@@

```{r statsassign, fig.align='center', fig.cap="Distribution of Year 1 GPA relative to Probation Cutoff for All Students", fig.width=6, fig.height=3, echo= FALSE, message=FALSE, warning=FALSE}
# Check if the RDD is "sharp" or "fuzzy"
ggplot(clean_cut,aes(x = dist_from_cut, y = status, group = status, color = status))+
  geom_point(alpha = 0.05)+
  geom_smooth(method = "lm")+
  theme_minimal()+
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))+
  labs( title = "Stuent Academic Status",
        y ="Academic Status",
        x = "Year 1 GPA minus Probation Cutoff")
```

@@

### Model Construction

@@

&nbsp; &nbsp; As the GPA is an continuous variable, I would use the linear regression to estimate the probation impact on student annual GPA in the second academic year. All of relevant model results will be discussed in the Results Section \@ref(result).

@@

## Impact on Dropout

&nbsp; &nbsp; The imme

```{r, include=FALSE}
# import data
clean_cut <- read.csv(here("inputs/data/bandwidth02.csv"))
clean_cut2 <- read.csv(here("inputs/data/bandwidth03.csv"))
clean_cut3<- read.csv(here("inputs/data/bandwidth06.csv"))
```

@@

# Bandwidth 0.3
dropout_03 <- glm(left_school ~ status + dist_from_cut + gender + highHS + first_language ,family = "binomial", data = clean_cut2)

# Bandwidth 0.6
dropout_06 <- glm(left_school ~ status + dist_from_cut + gender + highHS + first_language ,family = "binomial", data = clean_cut3)
```

@@

```{r dropfig, fig.align='center', fig.cap="Academic Probation Impact on Dropout", fig.width=8, fig.height=4, echo= FALSE, message=FALSE, warning=FALSE}
clean_cut$probation_year1 <- as.factor(clean_cut$probation_year1)
ggplot(data= clean_cut, aes(x = dist_from_cut, y = left_school, color = status))+
  geom_point(alpha = 0.05)+
  geom_smooth(method = "glm", method.args = list(family = 'binomial'))+
  theme_light()+
  theme(legend.title = element_blank(), legend.position = "bottom", plot.title = element_text(hjust = 0.5))+
  guides(color=guide_legend(override.aes=list(fill=NA)))+
  labs(title = "Academic Probation Impact on Dropout",
       x = "Year 1 GPA minus Probation Cutoff",
       y = "Left University")
```

## Impact on Subsequent Performance

```{r, include=FALSE}
# Linear Regression to estimate the impact of probation on student Year 2 GPA

# Bandwidth 0.2
gpa_02 <- lm(GPA_year2 ~ status + dist_from_cut + gender + highHS + first_language, data = clean_cut)
#summary(gpa_02)
# Bandwidth 0.3
gpa_03 <- lm(GPA_year2 ~ status + dist_from_cut + gender + highHS + first_language, data = clean_cut2)

# Bandwidth 0.6
gpa_06 <- lm(GPA_year2 ~ status + dist_from_cut + gender + highHS + first_language, data = clean_cut3)
```

@@

```{r gpafig, fig.align='center', fig.cap="Academic Probation Impact on Year 2 GPA", fig.width=8, fig.height=4, echo= FALSE, message=FALSE, warning=FALSE}
ggplot(data= clean_cut, aes(x = dist_from_cut, y = GPA_year2, color = status))+
  geom_point(alpha = 0.05)+
  geom_smooth(method = "lm")+
  theme_light()+
  theme(legend.title = element_blank(), legend.position = "bottom", plot.title = element_text(hjust = 0.5))+
  guides(color=guide_legend(override.aes=list(fill=NA)))+
  labs(title = "Academic Probation Impact on Year 2 GPA",
       x = "Year 1 GPA minus Probation Cutoff",
       y = "Year 2 GPA")
```

@@

@@

# Bandwidth 0.3
grad_03 <- glm(gradin4 ~ status + dist_from_cut + gender + highHS + first_language ,family = "binomial", data = clean_cut2)

# Bandwidth 0.6
grad_06 <- glm(gradin4 ~ status + dist_from_cut + gender + highHS + first_language ,family = "binomial", data = clean_cut3)
#summary(grad_06)
```

@@

```{r gradfig, fig.align='center', fig.cap="Academic Probation Impact on Graduation by Year 4", fig.width=6, fig.height=3, echo= FALSE, message=FALSE, warning=FALSE}
clean_cut$probation_year1 <- as.factor(clean_cut$probation_year1)
ggplot(data= clean_cut, aes(x = dist_from_cut, y = gradin4, color = status))+
  geom_point(alpha = 0.05)+
  geom_smooth(method = "glm", method.args = list(family = 'binomial'))+
  theme_light()+
  theme(legend.title = element_blank(), legend.position = "bottom", plot.title = element_text(hjust = 0.5))+
  guides(color=guide_legend(override.aes=list(fill=NA)))+
  labs(title = "Academic Probation Impact on Graduation by Year 4",
       x = "Year 1 GPA minus Probation Cutoff",
       y = "Graduate by Year 4")
```

# Discussion

## Summary

@@

>>>>

---
title: "Lack of Evidence?: Academic Probation Does Not Affect Student Dropout and Subsequent Performance"
subtitle: "Revisiting the Ability, Gender, and Performance Standards: Evidence from Academic Probation (Lindo, Sanders, and Oreopoulos 2010)"
author: "Hong Shi"
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: " Many universities use academic probation as a wake-up call to ensure currently enrolled students to achieve minimum academic standards. This paper examines the causal impact on first-year students' responses in a Canadian univesity after being placed on academic probation in a sharp regression discontinuity design. The results do not suggest that placing first year students on academic probation affects the dropout decision, subsequent annual GPA and graduation. While there are heterogeneous responses across students' highschool performance, gender, first language towards academic probation. #High Level Importance \\par\\textbf{Keywords:} Regression Discontinuity Design, Bandwidth, Academic Probation, Student Performance, P-value, Causality"
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
toc: FALSE

@@

&nbsp; &nbsp; This paper examines the causal impact on first-year university students after being placed on academic probation by exploiting the discontinuous probation GPA cutoff in a sharp regression discontinuity design (RDD). The data comes form a large Canadian university with three individual campuses. It is a replication of `“Ability, Gender, and Performance Standards: Evidence from Academic Probation."` [@cite_original] with adjustment on the bandwidth (i.e., student within certain grade points of the academic probation cutoff). After applying the adjusted bandwidth, the results are NOT consistent with Lindo's findings that being placed on probation at the end of the first year increases the student dropout, improves the GPAs of the remaining students and negatively affects graduation rates. While students with different gender, highschool academic performance or first language react differently to academic probation. Furthermore, I also find heterogeneous responses to academic probation across campus that is not mentioned in the original paper (Note: May adjust the result after running the model.).

&nbsp; &nbsp; The rest of this paper is organized as follows: In Data section (Section \@ref(Data)), I would first explain the domain background about the academic probation. Next, I introduce the data used in the original paper. And then, I would describe the Quasi-experiment design rationale and also raise concerns about the bandwidth used in the original paper. In Model section (Section \@ref(mainmodel)), I elaborate on the sharp RDD, address its assumptions and propose the model used in this study. In Results section (Section \@ref(result)), I show the impact of academic probation on student dropout, subsequent performance and graduation with the adjusted bandwidth. In Discussion section (Section \@ref(discuss)), I would first summarize the study findings, and then elaborate on the causality of the experiment, address external and internal validity and also Simpson's Paradox (also possibly Berkson's Paradox) regarding the campus variation. Finally, I conclude with some limitations and suggest future directions of this experiment. ^[Code and data are available at the GitHub repo: https://github.com/honn-ishinn/RDD_academic_probation.]

# Data {#Data}

&nbsp; &nbsp; The data comes from a large Canadian university with three individual campuses: one central campus (Campus 1) and two smaller satellite campuses (Campus 2 and Campus 3) from 1996 to 2005. I analyzed it using R [@citeR], and packages `tidyverse` [@cite_tidyverse],  `here` [@cite_here], `rdrobust` [@cite_rdrobust], `rdd`[@cite_rdd], `rdpower`[@cite_rdpower], `haven`[@cite_haven] . I used packages `bookdown` [@cite_bookdown], `kableExtra` [@cite_kableextra], , `finalfit` [@cite_finalfit], `modelsummary`[@cite_modelsummary], `broom` [@cite_broom] to format the document and referenced `Impact Evaluation in Practice` [@cite_impact] to evaluate this experiment.

@@

4. Restrict students having their academic standing evaluated at the end of Year 1 ($98\%$ of the remaining sample).
```{r, include=FALSE}
clean_data <- read.csv(here("inputs/data/cleaned_original.csv"))
```

@@

@@

&nbsp; &nbsp; In this study, since the RDD only provides the most accurate estimates around the probation cutoff where treatment and comparison are most similar [@cite_impact], it is critical to determine the bandwidth(i.e., students within certain grade points of the probation cutoff score) to maintain the balance of observed characteristics of students around the probation cutoff. In Lindo's original paper, they choose the bandwidth of 0.6 and the following table (Table \@ref(tab:lindo)) shows the summary statistics of student observable characteristics within 0.6 GPA.

@@

```{r lindo, tab.cap="Summary Statistics of Students Within 0.6 GPA of Probation Cutoff", echo = FALSE, message = FALSE, warning=FALSE}
kable(combine3[,c(4,1,2,3)],caption = "Summary Statistics of Students Within 0.6 GPA of Probation Cutoff",
       booktabs = TRUE, linesep = "") %>% 
  kable_styling(latex_options = "hold_position")
```
&nbsp; &nbsp; According to @cite_GPA_determinants, highschool grade is one of the key determinants of university academic success. Thus, highschool grades could somehow reveal students' study potential in the university, so it is preferable to balance student highschool performance to obtain a more accurate estimate of academic probation. However, the t-test (Table \@ref(tab:ttesthslin)) conducted on highschool grade percentile suggests a large difference on highschool performance between student observations in good standing and on academic probation within Lindo's 0.6 GPA bandwidth. In addition, the study on the course load on academic success [@cite_credits] implies that course credit taken by students could reflect student academic ability. Therefore, another t-test (Table \@ref(tab:ttestcrlin)) is conducted on total course credits attempted^[Failing a course will not earn course credit but count as an attempted credit] in Year 1 between students in good standing and on probation. The result suggests a difference in the course workload between good standing and probation students within Lindo's 0.6 GPA bandwidth. As a result, the bandwidth that Lindo used in the original paper may include students with different academic potentials, reducing the local estimation accuracy of placing first-year students on academic probation.

@@

&nbsp; &nbsp; The following table (Table \@ref(tab:gradingtheme)) shows the grading theme of this Canadian university^[This is the current grading theme of this university as of 2021. The grading theme between 1996 and 2005 might be different and could not be retrieved on the university website. While the current grading theme may still serve as a reference of this study]. By applying Lindo's bandwidth of 0.6 GPA, observations barely above the probation cutoff roughly covers students within 60-66 average grade percentage (1.5-2.1 GPA in Campus 1 and 2, 1.6-2.2 GPA in Campus 3), and observations barely below the probation cutoff roughly covers students within 53-59 average grade percentage (0.9-1.5 GPA in Campus 1 and 2, 1.0-1.6 GPA in Campus 3). By converting the grade point value into grade percentage, it might become more obvious that there exists academic performance difference between students in good standing with 66 average grade percentage and students on probation with 53 average grade percentage. So I would argue that the bandwidth used in Lindo's is too wide to maintain similar characteristics, particularly academic potentials, between the probation and good standing students.

@@

&nbsp; &nbsp; The power of a test is the probability that the test will reject the test null hypothesis to detect a correct alternative hypothesis [@cite_stats_tests]. In this study, as the length of the bandwidth increases, more students are included within the bandwidth. The increase in the number of student raises the power to detect the effect on placing students on probation so that the effect is less likely to happen by chance rather than by the probation cutoff rule. Therefore, I use the `rdpower`[@cite_rdpower] package to calculate the sufficient sample size required (i.e., effect size) to detect the probation effect on (1) Student dropout, (2) Second year GPA and (3) Graduation rate by Year 4. The result as shown below (Table \@ref(tab:powersample)) suggests that the bandwidth of 0.2 GPA with 4106 students (1745 on probation, 2361 in good standing) could raise sufficient statistical power. Further adjustment on the bandwidth might either unbalance student characteristics (Increase bandwidth) or decrease the power of the study (Decrease bandwidth). So I would apply the bandwidth of 0.2 GPA to examine the impact of academic probation in the following sessions.

@@

# Model {#mainmodel}

@@

&nbsp; &nbsp; The university policy ensures that all first year students below the probation cutoff will be placed into academic probation^[The original data shows university administrative errors in data reporting that falsely place students either above the cutoff on probation or below the cutoff in good standing (98 out of 44362 student observations). These dirty data are removed from this study.]. So the discontinuity around the cutoff is "sharp".

@@

```{r statsassign, fig.align='center', fig.cap="Distribution of Year 1 GPA relative to Probation Cutoff for All Students", fig.width=6, fig.height=3, include =FALSE}
# Check if the RDD is "sharp" or "fuzzy"
# Skip this graph (a little redundant)
ggplot(clean_cut,aes(x = dist_from_cut, y = status, group = status, color = status))+
  geom_point(alpha = 0.05)+
  geom_smooth(method = "lm")+
  theme_minimal()+
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))+
  labs( title = "Stuent Academic Status",
        y ="Academic Status",
        x = "Year 1 GPA minus Probation Cutoff")
```

@@

### Model Construction {#model}

@@

&nbsp; &nbsp; As the GPA is an continuous variable, I would use the linear regression to estimate the probation impact on student annual GPA in the second academic year.

@@

&nbsp; &nbsp; Surprisingly, after restricting student samples within 0.2 probation cutoff GPA, the model results do not suggest significant academic probation impact on any student outcomes of interest (dropout, Year 2 GPA and graduation). This is inconsistent with Lindo's findings that there is significant academic probation impact on all outcomes of interest within 0.6 probation cutoff GPA.

## Impact on Dropout {#baseline}

&nbsp; &nbsp; As mentioned in Model Section \@ref(model), logistic regression is used to estimate the probation impact on probability of student dropout.
$$Prob(Dropout_{i}) = \beta_{0}Proation_{i} + f({Cutoff}\:{Distance}_{i}) + \alpha_{1}Gender_{i}+\alpha_{2}HighSchool_{i}+\alpha_{3}Language_{i} +\epsilon_{i}$$

&nbsp; &nbsp; In this model, the baseline characteristics of the first year students is (1) on academic probation, (2) GPA distance from the probation cutoff is 0, (3) female, (4) highschool grade below the average, (5) first language is English. The following table (Table \@ref(tab:dropsummary)) shows regression results of student samples within different GPA bandwidth (0.2 GPA, 0.3 GPA, 0.6 GPA, respectively). Within 0.2 GPA bandwidth, compared to students on academic probation, students in good standing have 0.711 times odds to leave the university, i.e., $28.9\%$ decrease in the odds of leaving the university, while holding other baseline characteristics constant. **However**, the p-value of 0.223, which is greater than the widely defined 0.05 threshold, suggests that the first year dropout decision does not significantly differ between good standing students and probation students, holding other baseline characteristics constant. And this difference could also not be clearly detected around the probation cutoff as shown below (Figure \@ref(fig:dropfig)). While within 0.6 GPA bandwidth used in Lindo's original paper, the results do suggest significant difference that students in good standing have $34.8\%$ decrease in the odds of first year dropout, holding other baseline characteristics constant. Such difference is also significant within 0.3 GPA bandwidth.

&nbsp; &nbsp; Within 0.2 GPA bandwidth, there is significant difference on student dropout decision across first language: compared to English native speaking students, students whose first language are other languages have $54.5\%$ decrease in the odds of first year dropout, holding other baseline characteristics constant. No other significant difference is found on student gender and highschool performance.

```{r, include=FALSE}
# import data
clean_cut <- read.csv(here("inputs/data/bandwidth02.csv"))
clean_cut2 <- read.csv(here("inputs/data/bandwidth06.csv"))
clean_cut3<- read.csv(here("inputs/data/bandwidth03.csv"))
```

@@

# Bandwidth 0.3
dropout_03 <- glm(left_school ~ status + dist_from_cut + gender + highHS + first_language ,family = "binomial", data = clean_cut3)

# Bandwidth 0.6
dropout_06 <- glm(left_school ~ status + dist_from_cut + gender + highHS + first_language ,family = "binomial", data = clean_cut2)
```

@@

```{r dropfig, fig.align='center', fig.cap="Academic Probation Impact on Dropout", fig.width=8, fig.height=4, echo= FALSE, message=FALSE, warning=FALSE}
ggplot(data= clean_cut, aes(x = dist_from_cut, y = left_school, color = status))+
  geom_point(alpha = 0.05)+
  geom_smooth(method = "glm", method.args = list(family = 'binomial'))+
  theme_light()+
  geom_vline(xintercept = 0, linetype = "dotted")+
  theme(legend.title = element_blank(), legend.position = "bottom", plot.title = element_text(hjust = 0.5))+
  guides(color=guide_legend(override.aes=list(fill=NA)))+
  labs(title = "Academic Probation Impact on Dropout",
       x = "Year 1 GPA minus Probation Cutoff",
       y = "Left University")
```

## Impact on Second Year Annual GPA

&nbsp; &nbsp; As mentioned in Model Section \@ref(model), linear regression is used to estimate the probation impact on student second year annual GPA.
$$GPAYear2_{i} = \beta_{0}Proation_{i} + f({Cutoff}\:{Distance}_{i}) + \alpha_{1}Gender_{i}+\alpha_{2}HighSchool_{i}+\alpha_{3}Language_{i} +\epsilon_{i}$$

&nbsp; &nbsp; The model baseline characteristics are the same as mentioned in dropout Section \@ref(baseline). The following table (Table \@ref(tab:gpasummary)) shows regression results of student samples within different GPA bandwidth (0.2 GPA, 0.3 GPA, 0.6 GPA, respectively). Within 0.2 GPA bandwidth, compared to students on academic probation, students in good standing earn 0.055 less GPA than those on academic probation in the second year, while holding other baseline characteristics constant. **However**, the p-value of 0.245 suggests that the first year dropout decision does not significantly differ between good standing students and probation students, holding other baseline characteristics constant. And this difference could also not be clearly detected around the probation cutoff as shown below (Figure \@ref(fig:gpafig)). While within 0.6 GPA bandwidth used in Lindo's original paper, the results do suggest significant difference that students in good standing earn 0.145 GPA less than students on probation, holding other baseline characteristics constant. Such difference is also significant within 0.3 GPA bandwidth.

&nbsp; &nbsp; Within 0.2 GPA bandwidth, there is significant difference on student second year GPA across gender and highschool performance. Compared to female students, male students earn 0.120 less annual GPA in the second year, holding other baseline characteristics constant. Compared to students with relative poor highschool academic performance, students with satisfactory academic performance earn 0.099 more annual GPA in the second year, holding other baseline characteristics constant. No other significant difference is found on student first language.
```{r, include=FALSE}
# Linear Regression to estimate the impact of probation on student Year 2 GPA

# Bandwidth 0.2
gpa_02 <- lm(GPA_year2 ~ status + dist_from_cut + gender + highHS + first_language, data = clean_cut)
#summary(gpa_02)
# Bandwidth 0.3
gpa_03 <- lm(GPA_year2 ~ status + dist_from_cut + gender + highHS + first_language, data = clean_cut3)

# Bandwidth 0.6
gpa_06 <- lm(GPA_year2 ~ status + dist_from_cut + gender + highHS + first_language, data = clean_cut2)
```

@@

```{r gpafig, fig.align='center', fig.cap="Academic Probation Impact on Year 2 GPA", fig.width=8, fig.height=4, echo= FALSE, message=FALSE, warning=FALSE}
ggplot(data= clean_cut, aes(x = dist_from_cut, y = GPA_year2, color = status))+
  geom_point(alpha = 0.05)+
  geom_smooth(method = "lm")+
  theme_light()+
  geom_vline(xintercept = 0, linetype = "dotted")+
  theme(legend.title = element_blank(), legend.position = "bottom", plot.title = element_text(hjust = 0.5))+
  guides(color=guide_legend(override.aes=list(fill=NA)))+
  labs(title = "Academic Probation Impact on Year 2 GPA",
       x = "Year 1 GPA minus Probation Cutoff",
       y = "Year 2 GPA")
```

@@

&nbsp; &nbsp; As mentioned in Model Section \@ref(model), logistic regression is used to estimate the probation impact on probability of student graduating by Year 4.
$$Prob(Graduation_{i}) = \beta_{0}Proation_{i} + f({Cutoff}\:{Distance}_{i}) + \alpha_{1}Gender_{i}+\alpha_{2}HighSchool_{i}+\alpha_{3}Language_{i} +\epsilon_{i}$$
&nbsp; &nbsp; The model baseline characteristics are the same as mentioned in dropout Section \@ref(baseline). As shown in the following table (Table \@ref(tab:gradsummary)), within 0.2 GPA bandwidth, compared to students on academic probation, students in good standing have $26.8\%$ increase in the odds of graduating by Year 4, while holding other baseline characteristics constant. **However**, the p-value of 0.161 suggests that the graduation does not significantly differ between good standing students and probation students, holding other baseline characteristics constant. And this difference could also not be clearly detected around the probation cutoff as shown(Figure \@ref(fig:gradfig)). Within 0.3 GPA and 0.6 GPA bandwidth, the results also do not suggest difference on graduation between good standing students and probation students at 0.05 significance level, holding other baseline characteristics constant.

&nbsp; &nbsp; Within 0.2 GPA bandwidth, there is significant difference on student graduation across gender: compared to female students, male students have $49.9\%$ decrease in the odds of graduating by Year 4, holding other baseline characteristics constant. No other significant difference is found on student highschool performance and first language.

@@

# Bandwidth 0.3
grad_03 <- glm(gradin4 ~ status + dist_from_cut + gender + highHS + first_language ,family = "binomial", data = clean_cut3)

# Bandwidth 0.6
grad_06 <- glm(gradin4 ~ status + dist_from_cut + gender + highHS + first_language ,family = "binomial", data = clean_cut2)
#summary(grad_06)
```

@@

```{r gradfig, fig.align='center', fig.cap="Academic Probation Impact on Graduation by Year 4", fig.width=6, fig.height=3, echo= FALSE, message=FALSE, warning=FALSE}
clean_cut$probation_year1 <- as.factor(clean_cut$probation_year1)
ggplot(data= clean_cut, aes(x = dist_from_cut, y = gradin4, color = status))+
  geom_point(alpha = 0.05)+
  geom_smooth(method = "glm", method.args = list(family = 'binomial'))+
  theme_light()+
  geom_vline(xintercept = 0, linetype = "dotted")+
  theme(legend.title = element_blank(), legend.position = "bottom", plot.title = element_text(hjust = 0.5))+
  guides(color=guide_legend(override.aes=list(fill=NA)))+
  labs(title = "Academic Probation Impact on Graduation by Year 4",
       x = "Year 1 GPA minus Probation Cutoff",
       y = "Graduate by Year 4")
```

# Discussion {#discuss}

## P-value and RDD assumptions

&nbsp; &nbsp; Why does using student samples within the bandwidth of 0.6 GPA and even 0.3 GPA "has evidence" to suggest significant academic probation impact on student outcomes of interest (probability of immediate dropout, Year 2 annual GPA, probability of graduating by Year 4), while using student samples within the bandwidth of 0.2 GPA does not?

&nbsp; &nbsp; Since all of the result interpretations are based on p-values, reassessing the definition of p-value may help to explain the above inconsistent results. In simple common English, P-value means the probability that something checked wrong to happen. The lower p-value means a higher probability of something checked right to be true, and "is there an academic probation impact on first year student outcome" is the checkpoint in this study.

&nbsp; &nbsp; However, the p-value does not tell if the underlying assumptions of the checkpoint is correct or not. In the paper `Statistical tests, P values, confidence intervals, and power: a guide to misinterpretations`, Greenland defines the p-value in a general view "as a statistical summary of the compatibility between the observed data and what we would predict or expect to see if we knew the entire statistical model (all the assumptions used to compute the P value) were correct"[@cite_stats_tests]. Therefore, reviewing RDD rationales and assumptions helps to draw correct interpretations from p-values.

&nbsp; &nbsp; In this study, the regression discontinuity design estimates academic probation impact on first year students by calculating the difference between treatment students on probation barely below the cutoff and comparison students in good standing barely above the cutoff. Its estimate is the most accurate where the treatment and comparison students are most similar and this estimate may not be generalized to students whose first year GPA are further away from the probation cutoff [@cite_impact].

&nbsp; &nbsp; According to the student summary statistics (Table \@ref(tab:lindo)) and t-test conducted on students highschool performance (Table \@ref(tab:ttesthslin)) and credit attempted in the first year (Table \@ref(tab:ttestcrlin)) in Data Section \@ref(bandwidth), within the bandwidth of 0.6 GPA, students barely below the cutoff generally hold weaker academic potentials than those barely above the cutoff. And these academic differences also serves as key factors affecting student outcomes.

&nbsp; &nbsp; Even though Lindos's original paper did test whether students share similar observable characteristics around the probation cutoff GPA in Results Section A [@cite_original], it failed to access students characteristics that are further away from the cutoff. It is unlikely that a student with 1.0 GPA is as similar as one with 2.0 GPA, while they are all included within Lindo's bandwidth to estimate the impact of academic probation. As a result, using a wider bandwidth could violate the similarity assumption of RDD, and interpreting p-values from assumption violated models to seek "statistical significance" may become senseless.

```{r, include=FALSE}
library(scales)
  ggplot(clean_cut,aes(x = study_campus, fill = status))+
  geom_bar(aes(y = (..count..)/sum(..count..)), position = position_dodge())+
  scale_y_continuous(labels = scales::percent)+
  theme_minimal()
```

## Causal inference

&nbsp; &nbsp; So does placing first year students on academic probation affect student outcomes? By applying a narrower bandwidth of 0.2 GPA, student characteristics become more balanced (Table \@ref(tab:mycutoff)). Although the t-test still suggests a difference on highschool performance between treatment and comparison students (Table \@ref(tab:ttesthslin)), model results in Result Section \@ref(result) has already suggestted

@@

Within Province Data
